The value of early warning systems for pandemic response

### 2024-02-28 reproduce-results.qmd
creating calibration script

# Inter-jurisdiction mixing
We use the following approach to align a

@@

@@

<!--EWS systems provide meaningful value under base-case assumptions representing a hypothetical new pandemic with parameters mirroring those of SARS-Cov-2. -->
Assuming no NPIs interventions are put in place (No NPIs scenario), a new COVID-19-like pandemic would produce significant damage, causing `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "deaths_per_100k")` deaths per 100,000 people [mean (95% prediction interval of stochastic replications)] over a one-year time-frame (Table 1). The unmitigated pandemic would infect `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "epi_size")` percent of the population, causing `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH_illness")` dollars per person due to illness, and `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH_deaths")` dollars due to deaths. The total health cost would be projected at `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH")` dollars per person.

If policymakers follow the base-case NPI policy simulated in this paper, the death toll of the pandemic would be reduced from `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "deaths_per_100k")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "deaths_per_100k")` deaths per 100,000 population if no enhanced early-warning system (EWS) is adopted (NPIs w/o EWS scenario). As it was the case during the COVID-19 pandemic, such an achievement would involve a substantial public health effort. Under baseline assumptions, this effort would require `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L1plus_days")` days of public health interventions, of which `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L5_days")` days are spent under the most-restrictive policy intervention level. We project this effort to cost on the order of `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "CNPI")` per person. Based on those assumptions, the total cost of the pandemic would be reduced from `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "C")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "C")` per person.

```{r}
#| label: tbl-1
#| tbl-cap: Base-case results
gt::gt(r$table_1)
```

Notes: All costs are expressed as 2020 dollars per person. All outcomes were computed over a one-year time frame.

@@

```{r}
#| label: fig-1
#| fig-cap: Base-case results
#| column: screen
#| fig-format: svg
#| out-width: 100%
plot(r$fig_A_epi)
```

Other socio-economic and contextual variables also have a meaningful effect on the value of EWS for pandemic response, including NPI costs, NPI effectiveness, decision lags and the stringency with which decision-makers introduce nonpharmaceutical interventions (Supplementary Figure 1).

@@

Policy stringency (i.e., how stringent the case thresholds used to introduce or remove interventions) has a similar effect on the value of EWS relative to decision lags - the EWS provides the most value under the least desirable conditions for disease prevention. In a context where decision-makers are 50% more stringent than in our base-case assumptions (i.e., their case thresholds are half of its base-case), society is able to avert a substantial number of deaths, reducing mortality from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "deaths_per_100k", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 population without the EWS. In that scenario, the EWS yeilds `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "NMB", NMB_comparator == F)` dollars per person of net benefit since most of the deaths would already be prevented through more stringent NPIs. Conversely, lenient policy choises worsen overall outcomes but also imply a higher value of the enhanced surveillance system.

### NPI effectiveness
The effectiveness of NPIs to reduce transmission is a key unkonwn at the outset of a new pandemic. At our baseline parameterization, NPI effectiveness has a non-monotonic effect on the value of surveillance systems to pandemic response (Supplementary Figure X). If NPI effectiveness is 0%, an improved surveillance system has no value, as expected. The value of the enhanced surveillance system non-linearly increases with NPI effectiveness slowly for maximum NPI effectiveness below 50%, and peaked when the maximum NPI effectiveness reaches 60%, enough to reduce the baseline R0 of 2.5 below 1 at pandemic outset.  Beyond that point, increases in NPI effectiveness produce lower total costs, but reduce the value of the surveillance system since such betterFor a baseline R0 of 2.5, a maximum NPI effectiveness of 50% implies that even the maximum intervention level (i.e., a lockdown) is not able to reduce Rt below 1 at the outset of the pandemic.

### Policy stringency

the total cost of the pandemic is baseline o have a detrimental effect If decision-makers take two weeks to introduce and implement NPIs

### Decision lags and policy stringency

In our baseline assumptions, we assumed decision-makers take one week to update their risk classification levels based on the epidemiological data. That decision lag includes both the time it takes for public health decision-makers to

## Conditions for high-value pandemic response EW systems

{Describe conditions under which value is lower or higher}

Points from this section for discussion:

* Value of EWS hinges on:
  + timeliness of decision-making
  + adherence to NPIs
  + relative costs/severity of disease

>>>>

@@

## Early warning systems can shape disease and policy dynamics

**MAybe this is useful for a briefing or slides, but maybe not the paper**:
An early warning system that shifts the epidemiological time-series used to inform non-pharmaceutical decisions can change the course of disease when used with effective interventions (Figure 1). We consider a network of jurisdictions that are independently managed by their own public health departments. For simplicity, we consider two cases: One in which the jurisdictions base their decisions upon syndromic surveillance data, and another (counterfactual) scenario where they are able to use data from environmental sampling sources (i.e., without loss of generality, wastewater surveillance), which lead the conventional surveillance data by a few days. The two sets of jurisdictions are otherwise equal, and face the exact same pandemic.

@@

<!--EWS systems provide meaningful value under base-case assumptions representing a hypothetical new pandemic with parameters mirroring those of SARS-Cov-2. -->

Assuming no NPIs interventions are put in place (No NPIs scenario), a new COVID-19-like pandemic would produce significant damage, causing `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "deaths_per_100k")` deaths per 100,000 people [mean (95% prediction interval of stochastic replications)] over a one-year time-frame (Table 1). In the absence of any behavioral change, the unmitigated pandemic would infect `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "epi_size")` percent of the population, causing `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH_illness")` dollars per person due to illness, and `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH_deaths")` dollars due to deaths. The total health cost would be projected at `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH")` dollars per person.

If policymakers followed the base-case NPI policy simulated in this paper, the death toll of the pandemic would be reduced from `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "deaths_per_100k")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "deaths_per_100k")` deaths per 100,000 population if no enhanced early-warning system (EWS) is adopted (NPIs w/o EWS scenario). As it was the case during the COVID-19 pandemic, such an achievement would involve a substantial public health effort. Under baseline assumptions, this effort would require `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L1plus_days")` days of public health interventions, of which `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L5_days")` days are spent under the most-restrictive policy intervention level. We project this effort to cost on the order of `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "CNPI")` per person. Based on those assumptions, the total cost of the pandemic would be reduced from `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "C")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "C")` per person.

@@

### NPI Costs

NPI costs had a moderate effect on the value of surveillance systems. If NPI costs were 50% higher than assumed in our base-case scenario (i.e., if achieving the same NPI effectiveness were to cost 50% more than assumed under our base-case scenario), then the total cost of the pandemic would increase from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "C", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x NPI costs", variable == "C", NMB_comparator == T)` dollars per person with the conventional surveillance system. (Supplementary Figure 1, **Supplementary Table X**). Despite that increase, the NMB of the system would be only marginally reduced from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "NMB", NMB_comparator == F)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x NPI costs", variable == "NMB", NMB_comparator == F)` dollars per person, because the EWS system only marginally increased the number of days under interventions.

@@

Policy stringency (i.e., how stringent the case thresholds used to introduce or remove interventions) has a similar effect on the value of EWS relative to decision lags - the EWS provides the most value under the least desirable conditions for disease prevention. In a context where decision-makers are 50% more stringent than in our base-case assumptions (i.e., their case thresholds are half of its base-case), society is able to avert a substantial number of deaths, reducing mortality from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "deaths_per_100k", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 population without the EWS. In that scenario, the EWS yields `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "NMB", NMB_comparator == F)` dollars per person of net benefit since most of the deaths would already be prevented through more stringent NPIs. Conversely, if policy makers were more lenient policy choices would change the syst overall outcomes but also imply a higher value of the enhanced surveillance system.

### NPI effectiveness

NPI effectiveness (i.e., the % reduction in infectious contacts caused by NPIs) had a non-monotonic effect on the value of surveillance systems to pandemic response (**Supplementary Figure X**). If NPI effectiveness is 0%, an improved surveillance system has no value, as expected. The value of the enhanced surveillance system increases with NPI effectiveness slowly for maximum NPI effectiveness below around 50%, and peaked when the maximum NPI effectiveness reaches 60%, enough to reduce the baseline R0 of 2.5 below 1 at pandemic outset. In general, we find that the value of the EWS peaks at a maximum NPI effectiveness value at $1-1/R0$, for different values of R0. Beyond that point, increases in NPI effectiveness result in lower total pandemic costs, but the value of the surveillance system starts to decrease.
